A finite-state morphological transducer for Kyrgyz

### 2011-10-24 dev/paper/abstract.tex
more shortened forms

%\subsection{Voicing assimilation}\label{devoicing}
In Kyrgyz, there are two basic processes affecting the realisation of consonants when they are adjacent to other consonants: voicing assimilation and desonorisation.

Voicing assimilation in Kyrgyz involves the agreement of voicing of two consonants across a syllable boundary.  An example involves the locative and dative suffixes, as shown in table~\ref{exvoicing}.

\begin{table}[htbp]
	\caption{examples comparing voicing and devoicing in Kyrgyz}\label{exvoicing}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/алма-DA/ & [алмада] & \eng{apple--\gmk{LOC}} \\
			/каз-DA/ & [казда] & \eng{goose--\gmk{LOC}} \\
			/баш-DA/ & [башта] & \eng{head--\gmk{LOC}} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{2em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/алма-GA/ & [алмага] & \eng{apple--\gmk{DAT}} \\
			/каз-GA/ & [казга] & \eng{goose--\gmk{DAT}} \\
			/баш-GA/ & [башка] & \eng{head--\gmk{DAT}} \\
			\bottomrule
		\end{tabular}
	}
\end{table}

Here, the underlying /D/ is always realised as [д] when after a voiced segment (including a vowel), but is realised as [т] after an unvoiced consonant, while underlying /G/ is always realised as [г]\footnote{‹г› is pronounced [ɣ] in the context of front vowels and [ʁ] in the context of back vowels; it also has realisations of [ɡ] and [ɢ] after nasals.} after a voiced segment, but is realised as [к]\footnote{‹к› is realised as [q] in the context of back vowels and [k] in the presence of front vowels.} after voiceless consonants.

Voicing assimilation was dealt with simply by creating a single twol rule that transformed any relevant archiphoneme (\{B\}, \{G\}, \{D\}, \{L\}, and \{N\}) to a voiceless stop (either [п], [к], or [т]) after a voiceless consonant.  The set of voiceless consonants was also defined in twolc so that rules sensitive to this set would be easier to write.
%FIXME: and more efficient?  Or not?

%\subsection{Desonorisation}
Desonorisation happens to /N/ when it follows any consonant, and to /L/ when it follows a consonant of equal or lower sonority (i.e., /л/, nasals (/м/, /н/, /ң/), and obstruents, but not /й/, /р/, or vowels).  The resulting surface consonant is [д] for both /N/ and /L/, unless it follows a voiceless consonant, in which case it surfaces as [т].  An example includes the accusative and plural suffixes:

\begin{table}[htbp]
	\caption{examples comparing desonorisation of \{N\} and \{L\}}\label{exdesonorisation}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/алма-NI/ & [алманы] & \eng{apple--\gmk{ACC}} \\
			/сыр-NI/ & [сырды] & \eng{secret--\gmk{ACC}} \\
			/каз-NI/ & [казды] & \eng{goose--\gmk{ACC}} \\
			/баш-NI/ & [башты] & \eng{head--\gmk{ACC}} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{2em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/алма-LAр/ & [алмалар] & \eng{apple--\gmk{PL}} \\
			/сыр-LAр/ & [сырлар] & \eng{secret--\gmk{PL}} \\
			/каз-LAр/ & [каздар] & \eng{goose--\gmk{PL}} \\
			/баш-LAр/ & [баштар] & \eng{head--\gmk{PL}} \\
			\bottomrule
		\end{tabular}
	}
\end{table}

%FIXME: explain better
The orthography makes this straightforward, since e.g., words borrowed from Russian which end in ‹в› are treated in spoken Kyrgyz as unvoiced (and indeed have unvoiced surface forms syllable-finally), but are treated as voiced in the orthography.

Desonorisation was dealt with by a rule that changes \{L\} to [д] following a series of ``low-sonority'' consonants appropriately defined earlier in the twol file and a syllable boundary.  The \{N\} desonorisation rule changes \{N\} to [д] following a series containing all voiced consonants, followed by a syllable boundary.  Both sets were defined to exclude voiceless consonants that would instead trigger devoicing (§\ref{devoicing}).

%\subsection{Lenition}
In Kyrgyz, stem-final [voiceless] labial and dorsal consonants voice when a suffix beginning with a vowel follows them.  The implementation of this in twol is very straightforward.
%FIXME: what else needs to be said here?

%\subsection{Nouns ending in /рн/, etc.}

Kyrgyz has a number of nouns that underlyingly end with a consonant cluster consisting of two liquids, but which are split by an epenthetic vowel syllable-finally.
\begin{table}[htbp]
	\caption{examples of /рн/-final nouns surfacing with epenthetic vowel}\label{exrnvowel}
	\centering
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/мурн/ & [мурун] & \eng{nose} \\
			/мурн-LAр/ & [мурундар] & \eng{nose--\gmk{PL}} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{2em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/орн/ & [орун] & \eng{place} \\
			/орн-LAр/ & [орундар] & \eng{places--\gmk{PL}} \\
			\bottomrule
		\end{tabular}
	}
\end{table}
However, when a vowel-initial suffix follows the noun, the epenthetic vowel is absent, and all rules of consonant unfaithfulness involving consonant clusters are enforced[?].
%FIXME: "[?]"
%FIXME: make better references to the tables
\begin{table}[htbp]
	\caption{examples of /рн/-final nouns surfacing with no epenthetic vowel}\label{exrnnovowel}
	%\centering
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			%FIXME: should the glosses here and below be morpheme-by-morpheme for consistency, or is a more colloquial gloss okay?
			/мурн-(S)I/ & [мурду] & \eng{his/her/its nose} \\
			/мурн-Iм-DAн/ & [мурдуман] & \eng{from my nose} \\
			\bottomrule
		\end{tabular}
	}
	\hspace{0.5em}
	\subtable{
		\begin{tabular}{lll}
			\toprule
			underlying & surface & gloss \\
			\midrule
			/орн-(S)I/ & [орду] & \eng{his/her/its place} \\
			/орн-Iм-DAн/ & [ордуман] & \eng{from my place} \\
			\bottomrule
		\end{tabular}
	}
\end{table}

This was dealt with in a somewhat non-linguistic way: a default epenthetic vowel {y} was defined and inserted into the words by default:

%FIXME: do something with these: make them prettier, remove them, explain them better, ....?
\noindent орун:ор\%\{y\%\}н N-INFL ;\\
\noindent мунун:мур\%\{y\%\}н N-INFL ;

The vowel was then removed by a twol rule when a vowel followed.  To realise the epenthetic vowel, a rule resembling the \{I\} vowel harmony rule was created---the primary difference between these rules is that the \{y\} harmony rule only acts when something other than a vowel (i.e., a consonant or a word boundary) follows the following consonant.  While \{y\} will always behave like \{I\}, there is unfortunately no way to make this happen without replicating the context of the \{I\} rule.  This is due to restrictions in two-level morphology.
%FIXME: last sentence needs some rewriting

%\subsection{й+vowel letters}

In Kyrgyz, there are a series of letters which represent /й/ plus a vowel: /я, е, ё, ю/ represent /йа, йэ, йо, йу/ respectively.  However, /э/ is also represented as /е/ when short and after a consonant (i.e., it’s only represented as /э/ when long—/ээ/—or when word-initially).  These ``yoticised'' vowels proved difficult to work.  For example, /бой+(S)I/ \eng{length}, given normal vowel harmony rules, would surface as [бойу].  However, the correct form is [бою].  Distinct and separate vowel harmony rules had to be created for all vowel archiphonemes in post-vocalic contexts, e.g. here to turn \{I\} into [ю] after /й/.  An additional rule had to delete the underlying /й/ before yoticised vowel letters (so that e.g. [бойю] was not output).  Because there are too levels to the phonology in twolc and it’s otherwise not linear, attention had to be paid to the fact that e.g., the distinct vowel harmony rules for post-/й/ contexts recognised that the /й/ would not appear on the second [surface] level.

%\section{Statistics}

%FIXME: numbers!!
number of stems, number of continuation lexica, coverage, mean ambiguity (in corpus)

%\subsection{Number of stems}

Table~\ref{table:stems} gives the approximate number of stems in each main word class. The numbers
are approximate as some words may have two entries for one stem (for example if a set of forms
is irregular).

>>>>

%\subsection{Voicing assimilation}\label{devoicing}
In Kyrgyz, there are two basic processes affecting the realisation of consonants when they are adjacent to other consonants: voicing assimilation and desonorisation.  Voicing assimilation in Kyrgyz involves the agreement of voicing of two consonants across a syllable boundary.

%\subsection{Desonorisation}
Desonorisation involves the loss of sonority of sonorants in certain phonological environments (e.g., /n/ becomes [d]).

%\subsection{Lenition}
In Kyrgyz, stem-final [voiceless] labial and dorsal consonants voice when a suffix beginning with a vowel follows them.

%\subsection{й+vowel letters}
In Kyrgyz, there are a series of letters which represent /й/ plus a vowel: /я, е, ё, ю/ represent /йа, йэ, йо, йу/ respectively.  However, /э/ is also represented as /е/ when short and after a consonant (i.e., it’s only represented as /э/ when long—/ээ/—or when word-initially).  These ``yoticised'' vowels proved difficult to work with.  For example, /бой+(S)I/ \eng{length}, given normal vowel harmony rules, would surface as [бойу].  However, the correct form is [бою].
%\section{Statistics}

%\subsection{Number of stems}
Table~\ref{table:stems} gives the approximate number of stems in each main word class. The numbers are approximate as some words may have two entries for one stem (for example if a set of forms is irregular).
